What words ought to exist? \\ {\normalsize Coining with coinduction}

### 2011-03-12 trunk/wishlist/paper.tex
progress on paper, coin

My website ``{\sf snoot.org}'' has a number of games on it, including
a Scrabble clone called Scribble\footnote{{\tt
http://snoot.org/toys/scribble/}} and Boggle clone called
Muddle.\!\footnote{{\tt http://snoot.org/toys/muddle/}} This website
has been running for almost ten years, comprising over 150,000
Scribble games totaling 3.8 million words placed and 628,000 Muddle
games with over 10 million words found. During each game, players
repeatedly attempt to play words that aren't real. The computer
rebukes them, but hope really springs eternal with these people. It's
like they truly deeply wish to break out of the shackles of the
Official Scrabble Players Dictionary. So the first approach to
determining what words ought to exist is to analyze the words that
people tried to play, in order to try to extract the essence of
word-yearning.

This analysis is quite straightforward. I took the ten years of logs
files and extracted each attempt to play a word in Scribble or Muddle.
These log files are quite large, so the first step is just to get a
count, for each alleged word, and store those in a more convenient
format. There were XXX total words attemped in Scrabble and XXX in
Muddle. The most frequent ones appear in
Figure~\ref{fig:mostfrequent}. The most frequent words are all
legitimate words, since players have a bias towards attempting words
that will not be rebuked by the computer. Seeing the words that people
wish existed is a simple matter of filtering out the words that
already exist, using the Scrabble dictionary. XXX\% of the words
attempted in Scribble and XXX\% in Muddle were not real. The most
frequent ones appear in Figure~\ref{fig:mostfrequentfake}.

\begin{figure}
\caption{Someone needs to fill this in with the data.}
\label{fig:mostfrequent}
\end{figure}

\begin{figure}
\caption{Someone needs to fill this in with the data.}
\label{fig:mostfrequentfake}
\end{figure}

XXX I need to write an analysis of the list and put it here.

In my opinion these are pretty good candidates for words to exist.
Probably if you were playing someone really intense in Scrabble, and
he or she played one of these, and was super deadpan about it and
maybe had caused some crying contests before, and a known
sesquipedalianist, you would let these fly because they look like real
words to me. A point in their favor is that they are quite low-scoring
words in Scrabble; one might expect that the words most wished for are
like ``qzkwv'' for the scoring potential. The effect is probably due
to a few factors: Players are less likely to attempt obvious
non-words, common letters appear more often on the rack and on the
board and so the opportunity to play words like in
Figure~\ref{fig:mostfrequentfake} presents itself more frequently, and
in Muddle, there is no advantage to using unusual letters, except the
joy of being a weirdo. Nonetheless, this list is surely biased by the
specifics of Scribble and Muddle, and the question at hand is not just
what words ought to exist for the purpose of internet word games, but
for general purposes.

\begin{figure}
\caption{Someone needs a diagram or table here of the count histogram.}
\label{fig:gamedistribution}
\end{figure}

Another downside is that this method completely ignores the many words
that are attempted only once or a small number of times. The
distribution is heavy-tailed; XXX\% of the words attempted are not
even in the top XXX (Figure~\ref{fig:gamedistribution}). Though those
words individually are not good candidates to exist, like tiny stars
wished upon in the night sky,\!\footnote{Astronomers now agree that
stars do exist, by the way.} in aggregate they form a significant
planetarium that may tell us what {\it kind} of words people wish
existed. For example, if we saw that the words ``sweeeeeeet'',
``sweeeeeeeeeeeeet'', ``sweeeet'' and ``sweeeeeeeeeeeeeet'' occurred a
few times each, we could infer that people wished that words like
``sweet'' with strictly more than two {\it e}s were real words. They
might even be indifferent to the absolute number of {\it e}s, as long
as there existed some legal variation with more than two {\it e}s.
(This appears to be borne out by data. According to Google's
estimates, the words ``swe$^n$t'' for various medium-sized $n$
(10--20) appear on the Internet with similar frequency. The only
exception is ``sweeeeeeeeeeeeeeeeeeet'', with 19 {\it e}s, which
unexpectedly appears three times as often as 18 or 20 {\it e}s does.)
In order to lance these two boils, in the next section I explore
statistical methods for generalizing from lots of individual examples.

% XXX should really try to get this in the right column, since it
% overflows the margins
\begin{figure}
\begin{tabular}{r|r|l}
17,900,000 & 0 & swt \\
1,060,000 & 1 & swet \\
580,000,000 & 2 & sweet \\
1,310,000 & 3 & sweeet \\
806,000 & 4 & sweeeet \\
509,000 & 5 & sweeeeet \\
283,000$^1$ & 6 & sweeeeeet \\  % spell correction for five
170,000 & 7 & sweeeeeeet \\
115,000 & 8 & sweeeeeeeet \\
75,200 & 9 &  sweeeeeeeeet \\
94,300$^2$ & 10 & sweeeeeeeeeet \\ % spell correction for nine?
51,700 & 11 & sweeeeeeeeeeet \\
37,900 & 12 & sweeeeeeeeeeeet \\
32,000 & 13 & sweeeeeeeeeeeeet \\
25,300 & 14 & sweeeeeeeeeeeeeet \\
24,300 & 15 & sweeeeeeeeeeeeeeet \\
41,000$^3$ & 16 & sweeeeeeeeeeeeeeeet \\ % spell correction for 14
55,000 & 17 & sweeeeeeeeeeeeeeeeet \\
45,000 & 18 & sweeeeeeeeeeeeeeeeeet \\
133,000$^4$ & 19 & sweeeeeeeeeeeeeeeeeeet \\ % for 15
34,800 & 20 &  sweeeeeeeeeeeeeeeeeeeet \\
%
16,100$^5$ & 25 & sweeeeeeeeeeeeeeeeeeeeeeeeet \\ % for weeeeeeeeeeeeeeeeeeeeeeeee t
10,100 & 30 & sweeeeeeeeeeeeeeeeeeeeeeeeeeeeeet \\
2,800 & 40 &  sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
923 & 50 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
118 & 75 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
38 & 100 &    sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
?$^6$ & 200 &     sweeeeeeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\ % word is too long for Google
\end{tabular}
\caption{Frequency of ``swe$^n$t'' on the internet for various $n$, estimated
by Google. Notes: {\bf (1)} Spell correction offered for ``sweeeeet''. {\bf (2, 3, 4)} Spell corrections
offered to e$^{9}$, e$^{14}$ and e$^{15}$ respectively. {\bf (5)} Spell correction offered for ``weeeeeeeeeeeeeeeeeeeeeeeee t'' (?) {\bf (6)} With two hundred {\it e}s, the word is too long for Google, which asks me to ``try using a shorter word.'' Thanks Google, but I already did try the shorter ones.}
\label{ref:sweet}
\end{figure}

@@

The reason that people are more likely to play words like ``tane'' is
that the letters are common---they appear more often in words, and
more often in the Scrabble bag. But it's not simply a matter of the
frequency of letters; if it were, we would expect to see words like
``eee'' dominating the list, since {\it e} is the most common letter
in English.\!\footnote{Tied for first place with {\it n}, {\it g},
{\it l}, {\it i}, {\it s}, and {\it h}.} People do not play such words
often because they do not {\it seem} like real words. ``oilsoap''
seems more like a word than ``ioaopsl'' to most non-crazy people,
even though they contain the same letters. This is because we have
expectations on what letters are likely to appear next to one
another in words. This section is about modeling expectations on
what letters appear together, and then using that model to generate
the most likely words that don't yet exist.

@@

\begin{verbatim}
0.166666667 hel
0.166666667 helpful
0.166666667 hello
0.083333333 felpful
0.083333333 fello
0.083333333 fel
0.041666667 helafel
0.041666667 helafelpful
0.041666667 helafello
\end{verbatim}

% laplace smoothing?

@@

\section{Ockham's Regexp?}

>>>>

My website ``{\sf snoot.org}'' has a number of games on it, including
a Scrabble clone called Scribble\footnote{{\tt
    http://snoot.org/toys/scribble/}} and Boggle clone called
Muddle.\!\footnote{{\tt http://snoot.org/toys/muddle/}} This website
has been running for almost ten years, comprising over 150,000
Scribble games totaling 3.8 million words placed and 628,000 Muddle
games with over 10 million words found. During each game, players
repeatedly attempt to play words that aren't real. The computer
rebukes them, but hope really springs eternal with these people. It's
like they truly deeply wish to break out of the shackles of the
Official Scrabble Players Dictionary.\!\footnote{For the analyses in
  this paper that depend on a list of legal words, I actually use a
  modified version of SOWPODS, which is the tournament list used in
  Australia and the UK, and significantly more permissive than the US
  Tournament Word List. Though the modified version is non-canonical,
  I stuck with it because it's what's been in use on the site for
  ten years.} So the first approach to determining what
words ought to exist is to analyze the words that people tried to
play, in order to try to extract the essence of word-yearning.

This analysis is quite straightforward. I took the ten years of logs
files and extracted each attempt to play a word in Scribble or Muddle.
These log files are quite large, so the first step is just to get a
count, for each alleged word, and store those in a more convenient
format. There were 3,572,226 total words attempted\footnote{Here a
  word attempted is the major word of the play. This does not include
  incidental words (typically two-letter ones) formed in the
  perpendicular direction.} in Scribble and 13,727,511 in Muddle. The
most frequent ones appear in Figure~\ref{fig:mostfrequent}. Aside from
the one-letter ones, the most frequent words are legitimate words,
since players have a bias towards attempting words that will not be
rebuked by the computer.

Seeing the words that people wish existed is a simple matter of
filtering out the words that already exist, using the Scrabble
dictionary. (I also filtered out one-letter ``words''. It is easy to
see that no one-letter words should exist, again because of
ambiguities created in spelling bees. Not only when literally spelling
``bees'', but according to the official Scripps National Spelling Bee
rules, the speller may optionally pronounce the word to be spelled
before and after spelling it. So if ``s'' were a word, then the
following ridiculous exchange obtains: Judge: ``S. The letter {\it s}.
Etruscan origin.'' Speller: ``S. S. S.'' and the judge cannot tell if
the speller meant to state the word before and after, or thinks the
word is spelled ``sss''.) 22.3\% of the words attempted in Scribble
and 36.8\% in Muddle were not real. The most frequent ones appear in
Figure~\ref{fig:mostfrequentfake}.

\begin{figure}
\begin{tabular}{|rl@{\qquad}rl|}
\multicolumn{2}{l}{{\bf \large Scribble}} &
\multicolumn{2}{l}{{\bf \large Muddle}} \\
\hline
Count & Word & Count & Word \\
\hline
45,605  &  a        &     20,412  &  late  \\
42,315  &  i        &     19,405  &  rate  \\
32,499  &  d$^*$    &     19,276  &  dear  \\
12,981  &  in       &     19,049  &  tear  \\
12,851  &  oe       &     19,019  &  date  \\
12,528  &  s$^*$    &     18,771  &  lear  \\
12,207  &  re       &     18,423  &  deal  \\
11,159  &  tv       &     18,231  &  real  \\
10,720  &  jo       &     18,138  &  lead  \\
10,386  &  it       &     18,076  &  tale  \\
10,369  &  et       &     17,969  &  lane  \\
9,659   &  qua      &     17,956  &  sear  \\
9,218   &  xi       &     17,570  &  read  \\
9,099   &  go       &     17,193  &  teal  \\
9,052   &  ow       &     17,170  &  lean  \\
8,801   &  qat      &     17,071  &  dare  \\
8,602   &  aa       &     16,923  &  dale  \\
8,278   &  un       &     16,892  &  seal  \\
8,142   &  en       &     16,806  &  sale  \\
8,005   &  or       &     16,465  &  seat  \\
\hline
\end{tabular}
\caption{Most frequently attempted words in Scribble and Muddle. Asterisks
indicate non-words.}
\label{fig:mostfrequent}
\end{figure}

\begin{figure}

\begin{tabular}{|rl@{\qquad}rl|}
\multicolumn{2}{l}{{\bf \large Scribble}} &
\multicolumn{2}{l}{{\bf \large Muddle}} \\
\hline
Count & Word & Count & Word \\
\hline
11,159 &  tv         &     16,251 &  dane    \\
4,003  &  ok         &     6,156  &  rane    \\
2,862  &  iraq       &     5,603  &  sare    \\
2,725  &  zen        &     5,576  &  nate    \\
2,448  &  cho        &     4,863  &  mear    \\
1,538  &  viz        &     4,750  &  cale    \\
1,418  &  sdasda     &     4,616  &  nees    \\
1,396  &  von        &     4,568  &  nale    \\
1,136  &  etc        &     4,507  &  fale    \\
878    &  int        &     4,347  &  deat    \\
829    &  june       &     4,263  &  tean    \\
745    &  lp         &     4,251  &  nile    \\
719    &  zion       &     4,160  &  mens    \\
665    &  cia        &     4,087  &  deel    \\
661    &  jim        &     3,851  &  deam    \\
651    &  iraqi      &     3,828  &  dana    \\
648    &  ques       &     3,781  &  beed    \\
542    &  que        &     3,769  &  lans    \\
502    &  tim        &     3,725  &  tade    \\
\hline
\end{tabular}
\caption{Most frequently attempted non-words in Scrabble and Muddle.}
\label{fig:mostfrequentfake}
\end{figure}

There's a clear difference between these two lists. The Scribble list
is dominated by words involving difficult-to-play letters like {\it v}
(there are no legal two-letter {\it v}-words). Most of the words would
probably be acknowledged as real, just not legal in Scribble. The ones
that don't already have meanings, like ``cho'' and ``int'' and ``que''
seem to be pretty good candidates to exist. The Muddle list is all
four-letter words (the minimum allowed length) using common letters.
Other than the ones that are already words, like ``dane'' and ``nile''
and ``mens'' (as in ``mens section'' or ``the powerfuel weapon kills
hard so many mens''), these are all good candidates for words to
exist. Probably if you were playing someone really intense in
Scrabble, and he or she played one of these, and was super deadpan
about it and maybe had caused some crying contests before, and a known
sesquipedalianist, you would let these fly because they look like real
words to me. A point in their favor is that they would be quite
low-scoring words in Scrabble; not a {\it z} or {\it q} to be found.
Even in the Scribble list there's no ``qzkwv'' junk. The effect is
probably due to a few factors: Players are less likely to attempt
obvious non-words, common letters appear more often on the rack and on
the board and so the opportunity to play words like in
Figure~\ref{fig:mostfrequentfake} presents itself more frequently, and
in Muddle, there is no advantage to using unusual letters, except the
joy of being a weirdo. Nonetheless, these lists are surely biased by
the specifics of Scribble and Muddle, and the question at hand is not
just what words ought to exist for the purpose of internet word games,
but for general purposes.

\begin{figure}
\includegraphics[width=\linewidth]{wishlist-cdf}
\caption{Distribution of word frequency. Approximately 25,000
  different words (y axis) were issued 55 times or fewer (x axis).}
\label{fig:gamedistribution}
\end{figure}

Another downside is that this method completely ignores the many words
that are attempted only once or a small number of times. Players are
very creative; of the 564,610 unique words attempted, 501,939 of them
aren't real! The vast majority of words are attempted only a handful
of times (Figure~\ref{fig:gamedistribution}). Though those words
individually are not good candidates to exist, like tiny stars wished
upon in the night sky,\!\footnote{Astronomers now agree that stars do
  exist, by the way.} in aggregate they form a significant planetarium
that may tell us what {\it kind} of words people wish existed. For
example, if we saw that the words ``sweeeeeeet'',
``sweeeeeeeeeeeeet'', ``sweeeet'' and ``sweeeeeeeeeeeeeet'' occurred a
few times each, we could infer that people wished that words like
``sweet'' with strictly more than two {\it e}s were real words. They
might even be indifferent to the absolute number of {\it e}s, as long
as there existed some legal variation with more than two {\it e}s.
(This appears to be borne out by data. According to Google's
estimates, the words ``swe$^n$t'' for various medium-sized $n$
(10--20) appear on the Internet with similar frequency. The only
exception is ``sweeeeeeeeeeeeeeeeeeet'', with 19 {\it e}s, which
unexpectedly appears three times as often as 18 or 20 {\it e}s does;
see Figure~\ref{fig:sweet}.) In order to lance these two boils, in the
next section I explore statistical methods for generalizing from lots
of individual examples.

% XXX should really try to get this in the right column, since it
% overflows the margins
\begin{figure}
\begin{tabular}{r|r|l}
17,900,000 & 0 & swt \\
1,060,000 & 1 & swet \\
580,000,000 & 2 & sweet \\
1,310,000 & 3 & sweeet \\
806,000 & 4 & sweeeet \\
509,000 & 5 & sweeeeet \\
283,000$^1$ & 6 & sweeeeeet \\  % spell correction for five
170,000 & 7 & sweeeeeeet \\
115,000 & 8 & sweeeeeeeet \\
75,200 & 9 &  sweeeeeeeeet \\
94,300$^2$ & 10 & sweeeeeeeeeet \\ % spell correction for nine?
51,700 & 11 & sweeeeeeeeeeet \\
37,900 & 12 & sweeeeeeeeeeeet \\
32,000 & 13 & sweeeeeeeeeeeeet \\
25,300 & 14 & sweeeeeeeeeeeeeet \\
24,300 & 15 & sweeeeeeeeeeeeeeet \\
41,000$^3$ & 16 & sweeeeeeeeeeeeeeeet \\ % spell correction for 14
55,000 & 17 & sweeeeeeeeeeeeeeeeet \\
45,000 & 18 & sweeeeeeeeeeeeeeeeeet \\
133,000$^4$ & 19 & sweeeeeeeeeeeeeeeeeeet \\ % for 15
34,800 & 20 &  sweeeeeeeeeeeeeeeeeeeet \\
%
16,100$^5$ & 25 & sweeeeeeeeeeeeeeeeeeeeeeeeet \\ % for weeeeeeeeeeeeeeeeeeeeeeeee t
10,100 & 30 & sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
2,800 & 40 &  sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
923 & 50 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
118 & 75 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t \\
38 & 100 &    sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t  \\
?$^6$ & 200 & sweeeeeeeeeeeeeeeeeeeeeeeeee\ldots t     \\ % word is too long for Google
\end{tabular}
\caption{Frequency of ``swe$^n$t'' on the internet for various $n$, estimated
by Google. Notes: {\bf (1)} Spell correction offered for ``sweeeeet''. {\bf (2, 3, 4)} Spell corrections
offered to e$^{9}$, e$^{14}$ and e$^{15}$ respectively. {\bf (5)} Spell correction offered for ``weeeeeeeeeeeeeeeeeeeeeeeee t'' (?) {\bf (6)} With two hundred {\it e}s, the word is too long for Google, which asks me to ``try using a shorter word.'' Thanks Google, but I already did try the shorter ones.}
\label{fig:sweet}
\end{figure}

@@

The reason that people are more likely to play words like ``rane'' is
that the letters are common---they appear more often in words, and
more often in the Scrabble bag. But it's not simply a matter of the
frequency of letters; if it were, we would expect to see words like
``eee'' dominating the list, since {\it e} is the most common letter
in English.\!\footnote{Tied for first place with {\it n}, {\it g},
{\it l}, {\it i}, {\it s}, and {\it h}.} People do not play such words
often because they do not {\it seem} like real words. ``oilsoap''
seems more like a word than ``ioaopsl'' to most non-crazy people,
even though they contain the same letters. This is because we have
expectations on what letters are likely to appear next to one
another in words. This section is about modeling expectations on
what letters appear together, and then using that model to generate
the most likely words that don't yet exist.

@@

It basically makes sense to consider the probability of a whole word
to be the chance of simultaneously seeing each pair of letters in it,
which is just the product. So the word ``hel'' is $2/3$ (for {\bf
  \<h}) $\times$ 2/2 (for {\bf he}) $\times$ 4/4 (for {\bf el}) $\times$ 2/6 (for
{\bf l\>}), which is $0.222$. These are the most likely words overall (I
discuss how to generate such lists in Section~\ref{sec:coin}):

\vspace{0.2in}
\begin{tabular}{rl@{\quad}rl}
22.2\%  &  hel      &      2.5\%  &  helpfel  \\
11.1\%  &  helo     &      2.5\%  &  helafel  \\
 7.4\%  &  fel      &      1.9\%  &  fulo     \\
 3.7\%  &  hell     &      1.9\%  &  hello    \\
 3.7\%  &  felo     &      1.2\%  &  fell     \\
 3.7\%  &  ful      &      1.2\%  &  helafelo \\
\end{tabular}
\vspace{0.2in}

This is pretty good. These words resemble the ones we observed to build
the Markov chain, but are novel. I think helafelo is a pretty rad word,
right?

The next step is to build a Markov chain for a list of real words and
see what results.

% laplace smoothing?
% TODO: the empty string?

TODO: most likely words in italian, too.

@@

\subsection{Coining words with coinduction} \label{sec:coin}

% \section{Ockham's Regexp?}
